Host richness increases the occurrence but not the severity of bark beetle-induced tree mortality

### 2021-10-06 Code/Manuscript-SMOTE.Rmd
updated analyses

@@

Paragraph 1: Broad applied context

In this context there is a particular need to understand interactions between outbreaks of irruptive insects, which in recent years (2003-2012) have affected 85 million hectares of forest globally, or ca. 18 million hectares more than wildfire (van Lierop et al. 2015).

- insect outbreaks are causing widespread tree mortality in temperate forest worldwide
- future changes in climate are expected to increase tree mortality
- forests provide key ecosystem services
- management

Paragraph 2: Theoretical context - diversity & natural enemies

Many studies have sought to understand how community diversity influences interactions between natural enemies and their resources. In the case of plant-herbivore interactions, the greater plant diversity often reduces herbivory effects on plants at both the individual and population level [i.e. 'associational resistance, @barbosa_associational_2009, @jactel_tree_2021]. The resource concentration hypothesis suggests that specialists insect herbivores are less likely to find and stay in stands where their hosts are less abundant [@root_organization_1973]. This may occur if insects are less likely to find a host tree due [i.e. 'host dilution effect', @] or chemical masking of the focal tree by heterospecific neighbors [i.e. 'host apparency hypothesis', @Castagneryrol_effects_2014]

[i.e., 'resource concentration hypothesis', @root_organization_1973].

1 - associational resistance - the abundance or damage of herbivores is lower when plant diversity is greater [@barbosa_associational_2009, @jactel_tree_2021]
hypotheses:
1 - resource concentration: the presence of heterospecific neighbors around a tree of a focal species grown in mixed stands leads to a lower probability of host tree finding by insect herbivores due to lower host abundance or frequency [@root_organization_1973]
2 - host apparency: 
3 - decoy: preference for non-host trees

H
Most studies have focused on individual- or population-level outcomes.

For instance, increased host community diversity may either increase or decrease the susceptibility of individuals to their natural enemies.

At the population-scale these effects may either scale-up or result in counter-intuitive effects. However, little research has examined community-level outcomes, which are often hard to quantify because of the number of potential interactions increases dramatically with increasing community diversity, particularly when natural enemies are generalists. Further, the effects of different natural enemies on their focal host populations often differ greatly, due to differences in susceptibility and mortality rates and the availability and quality (as viewed by their natural enemies) of resource communities. Critically, community-level outcomes may drive ecosystem processes, particularly when community diversity is low and natural enemy-resource relationships are highly dynamic.

*** paragraph about irruptive species and community host diversity*** 
- stand scale
- landscape scale

To better understand how community diversity influences interactions between natural enemies and their resources, we use a natural system with inherently low resource and natural enemy diversity. Specifically our research focuses on tree mortality due to three bark beetle species, the mountain pine beetle (*Dendroctonus ponderosae*), spruce beetle (*D. rufipennis*), and western balsam bark beetle (*Dryocoetes confusus*), which in subalpine forests of Western North America predominantly attack lodgepole pine (*Pinus contorta*), Engelmann spruce (*Picea engelmannii*), and subalpine fir (*Abies lasiocarpa*), respectively. We use this system to ask whether host tree richness or identity influence the occurrence and severity of tree mortality due to bark beetles at the community-scale? We hypothesize that
greater host richness will increase the probability of outbreak occurrence

Bark beetles (Curculionidae: Scolytinae) are among the few native insect species that can kill large numbers of trees in a single year. Bark beetles bore through the bark, where they mate and oviposit their eggs. Concurrently, bark beetles introduce pathogenic fungi Larvae feeding upon the phloem and fungal spread stop the translocation of water and nutrients and cause tree death. Conifer defense against bark beetles consists primarily of resin exudation that physically expels the beetle and allelochemicals, which repel or kill beetles. To overcome these defenses and colonize live trees, bark beetles rely on a mass-attack strategy, where pioneering beetles emit aggregation pheromones that call conspecifics to the tree. Typically bark beetles exist at low population levels and attack weakened trees, but as populations increase bark beetles attack increasingly better defended trees.For instance, in the continental western United States bark beetles have killed more than XXXX trees over the past X decades. Such severe mortality occurs only during outbreaks when pheromone-mediated mass-attack allows bark beetles to overcome tree defenses.

@@

To test these hypotheses, we used a large dataset consisting of XXX,XXX plots established by the United States Forest Service Forest Inventory and Analysis Program (FIA; https://www.fia.fs.fed.us/).

@@

@@

### Model the suitable stand conditions
Construct random forest models using synthetic minority oversampling technique (SMOTE), an approach for dealing with imbalanced data. The SMOTE approach oversamples the minority class by synthesizing new cases from the minority class. We compared this approach with two other common approaches for dealing with imbalanced data in a random forest framework - balanced and weighted random forests and selected the approach that classified affected plots with the greatest accuracy.

@@

# Results
## Is co-occurrence of multiple hosts common?
```{r describe frequency of hosts and agents, eval=T}
## Hosts
hosts <- df[!df$focalhost=="Any",] %>% group_by(focalhost) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
hostgroups <- df[df$focalhost=="Any",] %>% group_by(hosts) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
hostgroups$hosts <- factor(hostgroups$hosts, levels=hostgroups$hosts, labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))

@@

ggsave( here("Results/Figures/FigureHostFreq.jpg"), plot=gghostgroups,   width=fig.width1,height=fig.width1*0.75,units="in")
```

Yes, 66% (n=5,850) FIA plots had at least two host species present. Subalpine fir was most likely to occur with other hosts (50% of stands), followed by Engelmann spruce (45% stands), and lodgepole pine (36% of stands) (Fig. \@ref(fig:FigHostFreq)).

```{r FigHostFreq, fig.cap = ".", out.width = "322.56pt"}
knitr::include_graphics( here("Results/Figures/FigureHostFreq.jpg"))
```

## Is co-occurrence of the stands conditions suitable for multiple bark beetle species common?

20% of stands were suitable for more than one bark beetle species. Conditions suitable for WBBB were most likely to co-occur with conditions suitable for another bark beetle (17% of stands)

@@

@@

ggsave( here("Results/Figures/FigureSuitable.jpg"), plot=ggfreq,width=fig.width1,height=fig.width1*0.75,units="in")
```

(Fig. \@ref(fig:FigureSuitable)).

```{r FigureSuitable, fig.cap = ".", out.width = "322.56pt"}
knitr::include_graphics( here("Results/Figures/FigureSuitable.jpg"))
```

@@

ggsave( here("Results/Figures/FigureOccurrencebySuitablehosts.jpg"), plot=ggoccurrence,  width=fig.width2, height=fig.width1*0.66,units="in")

@@

Given suitable stand conditions for multiple agents, the probability of occurrence  of any agent was significantly greater than the probability of a single agent \@ref(fig:FigureSuitable).
```{r FigOccurrence, fig.cap = "Severity",  out.width = "504.8pt"}
knitr::include_graphics( here("Results/Figures/FigureOccurrencebySuitablehosts.jpg"))
```

Given suitable stand conditions for multiple agents, is the severity of cumulative bark beetle-driven tree mortality greater in stands with multiple hosts?

In stands suitable for all three hosts, cumulative mortality was significantly greater than mortality by a single agent. \@ref(fig:FigSeverity). When stand conditions were suitable for only two bark beetles, the severity of cumulative tree mortality was driven by the most common agent (MPB > SB > WBBB) and was not significantly different than the cummualtive mortality.

@@

ggseverity <- ggplot(dfpred, aes(x=focalagent, y=propTBA*100, fill=focalagent, colour=focalagent))+geom_boxplot(notch=T)+facet_grid(.~suitablehostsP, scales="free_x", space="free_x")+xlab("bark beetle species")+scale_fill_manual(values=adjust_transparency(c(p1col[1:3], gray(0.3)), 0.5))+scale_colour_manual(values=c(p1col[1:3], gray(0.3)))+theme(legend.position = "none")+ylim(1,110)+geom_text(aes(label=pplot), vjust=-0.1,size=9, col="black")+ylab("Severity (% BA affected)")

ggsave(here("Results/Figures/FigureSevertitybySuitablehosts.jpg"), plot=ggseverity, width=fig.width2,height=fig.width1*0.66,units="in")

```{r FigSeverity, fig.cap = "Severity", out.width = "504.8pt"}
knitr::include_graphics( here("Results/Figures/FigureSevertitybySuitablehosts.jpg"))
```

## Given suitable stand conditions, is co-occurrence of multiple agents common?

Not really, only 21.3% (n=461) of stands suitable for multiple agents were affected by multiple agents. When multiple agents occur, the most commonly occurring agents was MPB (n=341), followed by SB (n=317) and WBBB (n=308). The most commonly occurring combinations of agents was MPB and SB (33.2% of cases; n=153), followed by MPB and WBBB (31.2% of cases; n=144) and then SB and WBBB (26% of case; n=120). The combination of all three agents was rare (9.5% of cases; n=44).

```{r determine co-occurrence of agents by hosts}
dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
dfpred <- dfpred[!dfpred$agentsuitable=="none",]

dfpred[dfpred$focalagent=="Any" &dfpred$suitablehosts %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB")& dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB") ,]  %>% group_by( agentsF) %>% summarise(n = n())%>% mutate(freq = n / sum(n))

@@

dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),]  %>% summarise(n = n())

MPBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "MPB+WBBB", "MPB+SB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
sum(MPBx$n)
sum(MPBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])

SBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+SB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
sum(SBx$n)
sum(SBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])

WBBBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PIEN\n&\nABLA", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
sum(WBBBx$n)
sum(WBBBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PIEN\n&\nABLA", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])

### 
table(dfpred$suitablehostsF, dfpred$agents)
x <- dfpred[dfpred$focalhost=="Any" & dfpred$suitablehostsF %in% c("PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") &  !dfpred$agentsFplongrev =="none",]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))

@@

co[co$agentsFplongrev %in% c("MPB, SB, & WBBB", "SB & WBBB", "MPB & WBBB", "MPB & SB") & co$suitablehostsF %in% c("PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),]

@@

##  Is severity greater in stands with multiple agents?
```{r multiple agents}
library(rcompanion)
dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
dfpred <- dfpred[!dfpred$suitablehosts == "none", ]
dfpred <- dfpred[dfpred$affected01=="affected",]

@@

Generally, yes.

@@

pBA1 <- ggplot(ppdat[ppdat$var=="BA",], aes(x=value, y=yhat, col=focalagent))+geom_smooth()+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )

pBA2 <- ggplot(xm[xm$var=="BA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("bark beetle")+xlab("basal area")

pQMD1 <- ggplot(ppdat[ppdat$var=="QMD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth()+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )

>>>>

Key findings

-

@@

Paragraph 1: Theoretical context - plant & insect interactions

Many studies have sought to understand how biodiversity influences interactions between natural enemies and their resources. In the case of plant-herbivore interactions, greater plant diversity often reduces herbivory on plants at both the individual and population level [i.e., 'associational resistance', @barbosa_associational_2009]. This is hypothesized to occur when insect herbivores, particularly specialists, are less likely to find and stay in areas where their hosts are less abundant [i.e., 'resource concentration hypothesis', @root_organization_1973]. This may occur if insects are less likely to encounter their host due to decrease in host frequency or abundance [i.e. 'host dilution effect', @], heterospecific neighbors chemically mask host trees [i.e. 'host apparency hypothesis', @castagneryrol_effects_2014], and/or mixed stands favor greater abundance of herbivore natural enemies [i.e., 'enemies hypothesis', @elton_ecology_1985, @staab_influence_2020].

#However, increased plant diversity may also result in an increase in herbivore richness ['resource specialization hypothesis', @keddy_plant_1984].

Yet little research has examined community-level outcomes, which are often hard to quantify because of the number of potential interactions increases dramatically with increasing community diversity, particularly when natural enemies are generalists. Further, the effects of different natural enemies on their focal host populations often differ greatly, due to differences in susceptibility and mortality rates and the availability and quality (as viewed by their natural enemies) of resource communities. Critically, community-level outcomes may drive ecosystem processes, particularly when community diversity is low and plant-herbivore relationships are highly dynamic. In this context there is a particular need to understand the effects of tree diversity on outbreaks of irruptive insects, which in recent years (2003-2012) have affected 85 million hectares of forest globally, or ca. 18 million hectares more than wildfire [@van_lierop_global_2015], resulting in widespread changes in carbon sequestration [e.g., @kurz_mountain_2008], timber production [e.g. @schowalter_insect_2012], and water quality [e.g., @Mikkelson_bark_2013], among other things.

Paragraph 2: Bark beetles

Bark beetles (Curculionidae: Scolytinae) are among the few native insect species that can kill large numbers of trees in a single year. For instance in the western United States, recent outbreaks of native bark beetles have killed 3.8 billion trees [1997-2018; @hicke_characterizing_2020]. Bark beetles bore through the bark, where they mate and oviposit their eggs. Concurrently, bark beetles introduce pathogenic fungi Larvae feeding upon the phloem and fungal spread stop the translocation of water and nutrients and cause tree death. Conifer defense against bark beetles consists primarily of resin exudation that physically expels the beetle and allelochemicals, which repel or kill beetles. To overcome these defenses and colonize live trees, bark beetles rely on a mass-attack strategy, where pioneering beetles emit aggregation pheromones that call conspecifics to the tree. Typically bark beetles exist at low population levels and attack weakened trees, but as populations increase bark beetles attack increasingly better defended trees. Such severe mortality occurs only during outbreaks when pheromone-mediated mass-attack allows bark beetles to overcome tree defenses.

@@

Paragraph 3: Bark beetles

To better understand how community diversity influences interactions between bark beeetles and their resources, we use a natural system with inherently low tree and bark beetle diversity. Specifically our research focuses on tree mortality due to three bark beetle species, the mountain pine beetle (*Dendroctonus ponderosae*), spruce beetle (*D. rufipennis*), and western balsam bark beetle (*Dryocoetes confusus*), which in subalpine forests of Western North America predominantly attack lodgepole pine (*Pinus contorta*), Engelmann spruce (*Picea engelmannii*), and subalpine fir (*Abies lasiocarpa*), respectively. We use this system to ask:

(1) Is co-occurrence of multiple trees susceptible to bark beetle attack common?
(2) How often do stand conditions susceptible to multiple bark beetles occur?
(3) Given stand stand conditions susceptible to multiple bark beetles, does the  the occurrence or severity of cumulative bark beetle mortality

@@

host tree richness or identity influence the occurrence and severity of tree mortality due to bark beetles at the community-scale?

To test these hypotheses, we used a large dataset consisting of XXX,XXX plots established by the United States Forest Service Forest Inventory and Analysis Program (FIA; https://www.fia.fs.fed.us/). Specifically we ask:

@@

### Model the suitable stand conditions

Construct random forest models using synthetic minority oversampling technique (SMOTE), an approach for dealing with imbalanced data. The SMOTE approach oversamples the minority class by synthesizing new cases from the minority class. We compared this approach with two other common approaches for dealing with imbalanced data in a random forest framework - balanced and weighted random forests and selected the approach that classified affected plots with the greatest accuracy.

@@

# Results

## Do stand conditions suitable for multiple outbreaks commonly co-occur?
```{r describe frequency of hosts and agents, eval=T}
## Hosts
hosts <- df[!df$focalhost=="Any",] %>% group_by(focalhost) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
hostgroups <- df[df$focalhost=="Any",] %>% group_by(hosts) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
hostgroups$hosts <- factor(hostgroups$hosts, levels=hostgroups$hosts, labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))

@@

@@

suitablehostgroups <- bbo.sub %>% group_by(suitablehostsP) %>% summarise(n = n())%>%mutate(freq = n / sum(n))

@@

ggsave( here("Results/Figures/FigureSuitable.jpg"), plot=ggfreq,width=fig.width1,height=fig.width1*0.75,units="in")

ggsave( here("Results/Figures/FigureHostFreq.jpg"), plot=gghostgroups,   width=fig.width1,height=fig.width1*0.75,units="in")

colnames(suitablehostgroups) <- colnames(hostgroups)
hostgroups$var <- "individual host"
suitablehostgroups$var <- "suitable host"

hosts.plot.dat <- rbind(hostgroups, suitablehostgroups)

gghostgroups <- ggplot(hosts.plot.dat, aes(x=hosts, y=n, fill=var))+geom_bar(stat="identity", position="dodge")+scale_fill_manual(values=c("grey", "black"))+ylab("frequency")+theme(legend.position="bottom")+theme(legend.title=element_blank(), axis.title.x=element_blank())
ggsave( here("Results/Figures/FigureHostCo.jpg"), plot=gghostgroups,   width=fig.width1, height=fig.width1*0.75, units="in")
```

Across the 10,439 FIA plots with one of the target host species, 66% (n=5,850) contained at least two of the target host species (Fig. \@ref(fig:FigHostFreq)). Subalpine fir was most likely to occur with other target host species (50% of plots), followed by Engelmann spruce (45% plots), and lodgepole pine (36% of stands). However, only 21% of all plots (n=2,163) were suitable for more than one bark beetle species. Conditions suitable for infestation in subalpine fir were most likely to co-occur with conditions suitable for infestation in another tree species (17% of stands).

```{r FigHostFreq, fig.cap = "The frequency of plots by host identity. Gray bars indicate host species were present (i.e., at least one individual of the focal species was present within the plot). Black bars indicate  stand structure and composition conditions were conducive to bark beetle infestation in focal tree species. For host identities, PICO is lodgepole pine, PIEN is Engelmann spruce, and ABLA is subalpine fir.", out.width = "244.8pt"}
knitr::include_graphics( here("Results/Figures/FigureHostCo.jpg"))
```

@@

ggsave( here("Results/Figures/FigureOccurrencebySuitablehosts.jpg"), plot=ggoccurrence,  width=fig.width2, height=fig.width1*0.5,units="in")

@@

Given suitable stand conditions for multiple agents, the probability of occurrence of bark beetle attributed mortality was significantly greater than the probability of a single bark beetle species \@ref(fig:FigureOccurrence).

```{r FigOccurrence, fig.cap = "The proportion of plots affected by the mountain pine beetle (MPB), spruce beetle (SB), western balsam bark beetle (WBBB), or any bark beetle (i.e., MPB, SB or WBBB) by the identity of hosts species susceptible to bark beetle infestation. For host identities, PICO is lodgepole pine, PIEN is Engelmann spruce, and ABLA is subalpine fir. Astericks above bars indiciate that proportion plots affected by the individual bark beelte species was significantly less than the proportion of plots affected by any bark beetle species.",  out.width = "504pt"}
knitr::include_graphics( here("Results/Figures/FigureOccurrencebySuitablehosts.jpg"))
```

## Is the severity of cumulative bark beetle-driven tree mortality greater in stands with multiple susceptible hosts?

@@

dfpred$focalagent2 <- factor(dfpred$focalagent, levels=levels(dfpred$focalagent), labels=c("MPB", "SB", "WBBB", "All"), ordered=T)

ggseverity <- ggplot(dfpred, aes(x=focalagent2, y=propTBA*100, fill=focalagent2, colour=focalagent))+geom_boxplot(notch=F)+facet_grid(.~suitablehostsP, scales="free_x", space="free_x")+xlab("bark beetle species")+scale_fill_manual(values=adjust_transparency(c(p1col[1:3], gray(0.3)), 0.5))+scale_colour_manual(values=c(p1col[1:3], gray(0.3)))+theme(legend.position = "none")+ylim(1,110)+geom_text(aes(label=pplot), vjust=-0.1,size=9, col="black")+ylab("Severity (% BA affected)")

ggsave(here("Results/Figures/FigureSevertitybySuitablehosts.jpg"), plot=ggseverity, width=fig.width2,height=fig.width1*0.66,units="in")
```

In stands susceptible for all three hosts, cumulative mortality was significantly greater than mortality by a single bark beetle species (Fig. \@ref(fig:FigSeverity). When stand conditions were suitable for only two bark beetles, the severity of cumulative tree mortality was driven by the most commonly occurring agent (MPB > SB > WBBB) and was not significantly different than the cumulative mortality (Fig. \@ref(fig:FigSeverity).

```{r FigSeverity, fig.cap= "Boxplots illustrating the severity of mortality (% of total basal area) attributed to the mountain pine beetle (MPB), spruce beetle (SB), western balsam bark beetle (WBBB), or all bark beetles (i.e., cumulative mortality due to MPB, SB and WBBB) by the identity of hosts species susceptible to bark beetle infestation. For host identities, PICO is lodgepole pine, PIEN is Engelmann spruce, and ABLA is subalpine fir. The bottom and top limits of each box are the lower and upper quartiles, respectively; the thick black line within the box is the median; error bars equal ±1.5 times the interquartile range; and points denote outliers, values outside ±1.5 times the interquartile range. Astericks above boxes indicate that severity bark beetle mortality attributed to the individual species was significantly less than the cumulative mortality due to all bark beetles.", out.width = "504pt"}
knitr::include_graphics( here("Results/Figures/FigureSevertitybySuitablehosts.jpg"))
```

## Is the  co-occurrence of multiple agents common?

Across all 10,439 FIA plots, only 5.3% (n=539) experience mortality due to multiple bark beetle species. However, MPB, SB, and WBBB were only present in 23.4% (n=2,441), 9.9% (n=1034), and 7.8% (n=813) of plots,respectively. Thus the number of plots with at least two bark beetle species co-occurring is more than 2x greater than would be expected if MPB, SB, WBBB occurred independently.

Given only the 2,163 stands where multiple tree species were susceptible to bark beetle infestation, 21.3% (n=461) were affected by multiple agents. When multiple tree species were susceptible to bark beetle infestation, the most commonly occurring agents was MPB (n=341), followed by SB (n=317) and WBBB (n=308). The most commonly occurring combinations of agents was MPB and SB (33.2% of cases; n=153), followed by MPB and WBBB (31.2% of cases; n=144) and SB and WBBB (26% of case; n=120). The combination of all three agents was extremely rare (9.5% of cases; n=44).

```{r determine co-occurrence of agents by hosts}
datBB.c <- df[df$focalhost=="Any",]
table(datBB.c$agents)
# nrow(datBB.c[datBB.c$agents %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB") ,])/nrow(datBB.c) 
# 5.3 % of plots were affected by multiple agents

table(datBB.c$agentSB)[2]/nrow(datBB.c) * table(datBB.c$agentMPB)[2]/nrow(datBB.c) * table(datBB.c$agentWBBB)[2]/nrow(datBB.c)

table(datBB.c$agentSB)[2]/nrow(datBB.c) * table(datBB.c$agentMPB)[2]/nrow(datBB.c)

datBB.c[datBB.c$agents %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB") ,]  %>% group_by( agents) %>% summarise(n = n())%>% mutate(freq = n / sum(n))

table(datBB.c$agentSB)

dfpred <- merge(BBpred, datBB.c, id.vars="PLT_CN", all=T)
dfpred <- dfpred[!dfpred$agentsuitable=="none",]

#dfpred[dfpred$focalagent=="Any" &dfpred$suitablehosts %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB")& dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB") ,]  %>% group_by( agentsF) %>% summarise(n = n())%>% mutate(freq = n / sum(n))

@@

#dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),]  %>% summarise(n = n())

MPBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "MPB+WBBB", "MPB+SB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
#sum(MPBx$n)
#sum(MPBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])

#sum(MPBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])

SBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+SB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
#sum(SBx$n)
#sum(SBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])

WBBBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PIEN\n&\nABLA", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
#sum(WBBBx$n)
#sum(WBBBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PIEN\n&\nABLA", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])

### 
#table(dfpred$suitablehostsF, dfpred$agents)
x <- dfpred[dfpred$focalhost=="Any" & dfpred$suitablehostsF %in% c("PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") &  !dfpred$agentsFplongrev =="none",]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))

@@

#co[co$agentsFplongrev %in% c("MPB, SB, & WBBB", "SB & WBBB", "MPB & WBBB", "MPB & SB") & co$suitablehostsF %in% c("PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),]

```{r determine co-occurrence of agents by suitable hosts}
dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
dfpred <- dfpred[!dfpred$agentsuitable=="none",]

#dfpred[dfpred$focalagent=="Any" &dfpred$suitablehosts %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB")& dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB") ,]  %>% group_by( agentsF) %>% summarise(n = n())%>% mutate(freq = n / sum(n))

#dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),]  %>% summarise(n = n())

MPBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "MPB+WBBB", "MPB+SB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
#sum(MPBx$n)
#sum(MPBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])

#sum(MPBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])

SBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+SB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
#sum(SBx$n)
#sum(SBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])

WBBBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PIEN\n&\nABLA", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
#sum(WBBBx$n)
#sum(WBBBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PIEN\n&\nABLA", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])

### 
#table(dfpred$suitablehostsF, dfpred$agents)
x <- dfpred[dfpred$focalhost=="Any" & dfpred$suitablehostsF %in% c("PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") &  !dfpred$agentsFplongrev =="none",]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))

#co[co$agentsFplongrev %in% c("MPB, SB, & WBBB", "SB & WBBB", "MPB & WBBB", "MPB & SB") & co$suitablehostsF %in% c("PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),]

@@

##  Is severity greater in stands with multiple agents?
```{r multiple agents, eval=F}
library(rcompanion)
dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
dfpred <- dfpred[!dfpred$suitablehosts == "none", ]
dfpred <- dfpred[dfpred$affected01=="affected",]

@@

@@

pBA1 <- ggplot(ppdat[ppdat$var=="BA",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )

pBA2 <- ggplot(xm[xm$var=="BA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot(se=F)+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("bark beetle")+xlab("basal area")

pQMD1 <- ggplot(ppdat[ppdat$var=="QMD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
